---
title: "When do voters accept democratic backsliding?"
subtitle: "A conjoint analysis of distinct types of backsliding on diverse dimensions of democracy"
author: "M. Vuković, J. Rovny, B. Stanley, P. Zagórski"
date: "07/14/2025"
date-format: long
slide-number: true
show-slide-number: all
format:
  revealjs: 
    css: presentation_style.scss
toc: true
toc-depth: 1
toc-title: Sections
logo: swps.jpg
---

## When do citizens oppose democracy?

-   Increased success of politicians defying liberal democracy
-   Puzzle: backsliding occurs despite declared support for democracy
-   Answers:
    -   High polarisation
    -   Trade-offs: disengaged and far-right voters less likely to punish undemocratic politicians
    -   Policy congruence diminishes propensity to punish

## Distinct transgressions

-   Different conceptions of democracy = different likelihood to punish undemocratic behavior
-   Populism, Illiberalism, Authoritarianism (PIA): distinct projects of backsliding
-   **H1a:** *Parties with liberal democratic features will be more likely to be chosen than parties with populist, illiberal or authoritarian (PIA) features.*

## Populism

-   Divides society into “the people” vs. “the elite”
-   Frames democracy as majoritarian, undermining minority protections
-   Dilutes democracy subtly, framed as a complement to democracy rather than an alternative (realisation of the "redemptive face")
-   **H1b:** *Populist violations of democracy are less likely to be punished by voters than illiberal or authoritarian violations.*

## Illiberalism

-   Practices and ideas against liberal democratic norms and values
-   Attacks personal freedom, pluralism, tolerance, dissent
-   More explicitly anti-liberal-democracy than populism, less extremely so than authoritarianism
-   **H1c:** *Illiberal violations of democracy are more likely to be punished by voters than populist violations.*

## Authoritarianism

-   Deference to authorities, subordination to collective, order over other values
-   Overtly rejects liberal democracy, elevates strong leader above the law
-   Suppresses opposition, dismantles oversight, may advocate violence
-   **H1d:** *Authoritarian violations of democracy will be the most strongly punished by voters compared to populist and illiberal violations.*

## Dimensions of democracy

-   Electoral (elections) and liberal (rights, rule of law) components
-   Voters care more about electoral features than liberal-democratic aspects
-   **H2:** *PIA violations of electoral democracy are more likely to be punished than violations of other democratic principles.*

## Policy congruence

-   Policy congruence reduces willingness to punish undemocratic behaviour
-   We examine how economic and cultural congruence interact with types of transgression
-   **H3:** *Citizens will be less likely to punish parties exhibiting PIA features when their policy preferences are congruent with those of the party.*

## Empirical strategy

-   Conjoint experiment in France, Italy, UK, Austria, Hungary, Poland, Czechia (N=7,000)
-   Three tasks, three rounds: compare hypothetical party profiles
-   Profiles vary by economic/cultural policy and one democratic dimension
-   Measures: AMCEs, forced choice, congruence by taxes/redistribution (economic) and migration (cultural)

## 

::: {style="font-size: 0.6em;"}
| Task | Dimension | Levels | Wording: We will... |
|----|----|----|----|
| ALL | Economic | Left-wing | Introduce high personal income taxes, collecting more from rich to fund a generous welfare state (e.g. health and educational public services). |
|  |  | Centrist | Introduce moderate personal income taxes, with the rich paying somewhat more to fund a reasonable welfare state. |
|  |  | Right-wing | Introduce low personal income taxes, with everybody paying the same rate to fund a minimal welfare state. |
| ALL | Cultural | Liberal | Accept migrants regardless of the country of origin. |
|  |  | Centrist | Accept only migrants that can fill important needs in our economy. |
|  |  | Conservative | Not accept any migrants no matter where they come from. |
:::

## 

::: {style="font-size: 0.5em;"}
| Task | Dimension | Level | Wording: We will... |
|----|----|----|----|
| 1 | Civil liberties | Populist | Restrict the funding of political organisations that explicitly oppose the will of the people. |
|  |  | Illiberal | Restrict the funding of political organisations that place the interests of small minorities over the common goals of our country. |
|  |  | Authoritarian | Ban political organisations that oppose the policies of our government. |
|  |  | Liberal (Ref.) | Ban political organisations only if they incite violence. |
| 2 | Rule of law | Populist | Fulfill the will of the majority of the people, even if it violates some existing law. |
|  |  | Illiberal | Contend that a law that does not serve the nation is illegitimate. |
|  |  | Authoritarian | Remove from office those who would prevent us from renewing our country. |
|  |  | Liberal (Ref.) | Respect all court rulings even if they go against the implementation of our program. |
| 3 | Fair elections | Populist | Ensure that electoral rules allow the voice of ordinary people to prevail over political elites. |
|  |  | Illiberal | Ensure that electoral rules do not give power to minorities to prevent us from realising our plans. |
|  |  | Authoritarian | Ensure that electoral rules allow for a strong leader that does not have to worry about the opposition. |
|  |  | Liberal (Ref.) | Ensure that electoral rules give all citizens an equal voice. |
:::

## 

```{r, echo=FALSE, warning=FALSE, fig.width=12, fig.height=6}
library(ggplot2)
library(dplyr)
library(tibble)

# Define the custom theme function provided
theme_plots <- function(base_size = 20, base_family = "Jost Regular") {
  theme_bw(base_size, base_family) +
    theme(
      panel.background = element_rect(fill = "#ffffff", colour = NA),
      title = element_text(size = rel(1), family = "Jost Bold", face = "plain"),
      plot.subtitle = element_text(
        size = rel(0.8),
        family = "Jost Regular",
        face = "plain"
      ),
      plot.caption = element_text(
        margin = margin(t = 10),
        size = rel(0.6),
        family = "Jost Light",
        face = "plain"
      ),
      panel.border = element_rect(
        color = "grey50",
        fill = NA,
        linewidth = 0.15
      ),
      panel.spacing = unit(1, "lines"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.25, colour = "grey90"),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_text(
        size = rel(0.8),
        family = "Jost Medium",
        face = "plain"
      ),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(hjust = 1, margin = margin(r = 10)),
      legend.position = "bottom",
      legend.title = element_text(
        size = rel(0.8),
        vjust = 0.5,
        family = "Jost Medium",
        face = "plain"
      ),
      legend.key.size = unit(0.7, "line"),
      legend.key = element_blank(),
      legend.spacing = unit(0.1, "lines"),
      legend.justification = "left",
      legend.margin = margin(t = -5, b = 0, l = 0, r = 0),
      strip.text = element_text(
        size = rel(0.9),
        hjust = 0,
        family = "Jost Regular",
        face = "plain"
      ),
      strip.background = element_rect(fill = "white", colour = NA),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")
    )
}


# Original data for violations
figure1_violations_data <- tribble(
  ~dimension                , ~party_type     , ~amce  , ~lower_ci , ~upper_ci ,
  "Civil liberties"         , "Populist"      ,  0.02  ,  0.01     ,  0.03     ,
  "Civil liberties"         , "Illiberal"     , -0.04  , -0.05     , -0.03     ,
  "Civil liberties"         , "Authoritarian" , -0.145 , -0.155    , -0.135    ,
  "Rule of law"             , "Populist"      , -0.06  , -0.07     , -0.05     ,
  "Rule of law"             , "Illiberal"     , -0.05  , -0.06     , -0.04     ,
  "Rule of law"             , "Authoritarian" , -0.075 , -0.085    , -0.065    ,
  "Free and fair elections" , "Populist"      , -0.04  , -0.05     , -0.03     ,
  "Free and fair elections" , "Illiberal"     , -0.16  , -0.17     , -0.15     ,
  "Free and fair elections" , "Authoritarian" , -0.17  , -0.18     , -0.16
)

# Add the 'Liberal' baseline data for plotting
# The AMCE and confidence intervals are all zero for the baseline
figure1_liberal_baseline <- tribble(
  ~dimension                , ~party_type , ~amce , ~lower_ci , ~upper_ci ,
  "Civil liberties"         , "Liberal"   ,     0 ,         0 ,         0 ,
  "Rule of law"             , "Liberal"   ,     0 ,         0 ,         0 ,
  "Free and fair elections" , "Liberal"   ,     0 ,         0 ,         0
)

# Combine the datasets
plot_data <- bind_rows(figure1_violations_data, figure1_liberal_baseline)

# Define the custom color palette as a named vector
party_colors <- c(
  "Authoritarian" = "red",
  "Illiberal" = "purple",
  "Populist" = "blue",
  "Liberal" = "orange"
)

# To ensure a logical order in the plot, convert 'party_type' to a factor
plot_data$party_type <- factor(
  plot_data$party_type,
  levels = c("Authoritarian", "Illiberal", "Populist", "Liberal")
)


# Create the final plot
amce_plot <- ggplot(
  plot_data,
  aes(
    y = dimension,
    x = amce,
    xmin = lower_ci,
    xmax = upper_ci,
    color = party_type
  )
) +
  # Add the points and error bars, dodged by party type
  geom_pointrange(position = position_dodge(width = 0.7), size = 0.9) +
  # Add a vertical line at zero for the baseline
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  # Apply the custom color scale
  scale_color_manual(values = party_colors) +
  # Apply the custom theme
  theme_plots() +
  # Set labels and titles
  labs(
    title = "AMCEs of deviations from liberal democracy",
    y = "Democratic dimension",
    x = "AMCE (Average Marginal Component Effect)",
    color = "Party type",
    caption = "Note: The zero line represents the 'Liberal' baseline."
  ) +
  # Reverse the y-axis to match the original paper's order
  scale_y_discrete(limits = rev)

# Print the plot
print(amce_plot)
```

## 

```{r, echo=FALSE, warning=FALSE, fig.width=12, fig.height=6}

library(ggplot2)
library(dplyr)

theme_plots <- function(base_size = 20, base_family = "Jost Regular") {
  theme_bw(base_size, base_family) +
    theme(
      panel.background = element_rect(fill = "#ffffff", colour = NA),
      title = element_text(size = rel(1), family = "Jost Bold", face = "plain"),
      plot.subtitle = element_text(
        size = rel(0.8),
        family = "Jost Regular",
        face = "plain"
      ),
      plot.caption = element_text(
        margin = margin(t = 10),
        size = rel(0.6),
        family = "Jost Light",
        face = "plain"
      ),
      panel.border = element_rect(
        color = "grey50",
        fill = NA,
        linewidth = 0.15
      ),
      panel.spacing = unit(1, "lines"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.25, colour = "grey90"),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_text(
        size = rel(0.8),
        family = "Jost Medium",
        face = "plain"
      ),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(hjust = 1, margin = margin(r = 10)),
      legend.position = "bottom",
      legend.title = element_text(
        size = rel(0.8),
        vjust = 0.5,
        family = "Jost Medium",
        face = "plain"
      ),
      legend.key.size = unit(0.7, "line"),
      legend.key = element_blank(),
      legend.spacing = unit(0.1, "lines"),
      legend.justification = "left",
      legend.margin = margin(t = -5, b = 0, l = 0, r = 0),
      strip.text = element_text(
        size = rel(0.9),
        hjust = 0,
        family = "Jost Regular",
        face = "plain"
      ),
      strip.background = element_rect(fill = "white", colour = NA),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")
    )
}

party_colors <- c(
  "Authoritarian" = "red",
  "Illiberal" = "purple",
  "Populist" = "blue",
  "Liberal" = "orange"
)

figure2_data <- tribble(
  ~dimension                , ~congruence     , ~party_type     , ~predicted_support , ~lower_ci , ~upper_ci ,
  # Data for Civil Liberties
  "Civil liberties"         , "Not congruent" , "Populist"      , 0.53               , 0.51      , 0.55      ,
  "Civil liberties"         , "Not congruent" , "Illiberal"     , 0.47               , 0.45      , 0.49      ,
  "Civil liberties"         , "Not congruent" , "Authoritarian" , 0.36               , 0.34      , 0.38      ,
  "Civil liberties"         , "Not congruent" , "Liberal"       , 0.52               , 0.50      , 0.54      ,
  "Civil liberties"         , "Congruent"     , "Populist"      , 0.63               , 0.61      , 0.65      ,
  "Civil liberties"         , "Congruent"     , "Illiberal"     , 0.58               , 0.56      , 0.60      ,
  "Civil liberties"         , "Congruent"     , "Authoritarian" , 0.47               , 0.45      , 0.49      ,
  "Civil liberties"         , "Congruent"     , "Liberal"       , 0.61               , 0.59      , 0.63      ,

  # Data for Rule of Law
  "Rule of law"             , "Not congruent" , "Populist"      , 0.45               , 0.43      , 0.47      ,
  "Rule of law"             , "Not congruent" , "Illiberal"     , 0.44               , 0.42      , 0.46      ,
  "Rule of law"             , "Not congruent" , "Authoritarian" , 0.42               , 0.40      , 0.44      ,
  "Rule of law"             , "Not congruent" , "Liberal"       , 0.51               , 0.49      , 0.53      ,
  "Rule of law"             , "Congruent"     , "Populist"      , 0.56               , 0.54      , 0.58      ,
  "Rule of law"             , "Congruent"     , "Illiberal"     , 0.57               , 0.55      , 0.59      ,
  "Rule of law"             , "Congruent"     , "Authoritarian" , 0.55               , 0.53      , 0.57      ,
  "Rule of law"             , "Congruent"     , "Liberal"       , 0.63               , 0.61      , 0.65      ,

  # Data for Elections
  "Free and fair elections" , "Not congruent" , "Populist"      , 0.52               , 0.50      , 0.52      ,
  "Free and fair elections" , "Not congruent" , "Illiberal"     , 0.40               , 0.38      , 0.42      ,
  "Free and fair elections" , "Not congruent" , "Authoritarian" , 0.39               , 0.37      , 0.41      ,
  "Free and fair elections" , "Not congruent" , "Liberal"       , 0.56               , 0.54      , 0.58      ,
  "Free and fair elections" , "Congruent"     , "Populist"      , 0.62               , 0.60      , 0.64      ,
  "Free and fair elections" , "Congruent"     , "Illiberal"     , 0.50               , 0.48      , 0.52      ,
  "Free and fair elections" , "Congruent"     , "Authoritarian" , 0.49               , 0.47      , 0.51      ,
  "Free and fair elections" , "Congruent"     , "Liberal"       , 0.67               , 0.65      , 0.69
)

figure2_data$party_type <- factor(
  figure2_data$party_type,
  levels = c("Authoritarian", "Illiberal", "Populist", "Liberal")
)

cultural_plot <- ggplot(
  figure2_data,
  aes(x = congruence, y = predicted_support, color = party_type)
) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  facet_wrap(~dimension, ncol = 3) +
  scale_color_manual(values = party_colors) +
  labs(
    title = "Effects of cultural policy congruence",
    x = "Cultural congruence",
    y = "Predicted support",
    color = "Party type"
  ) +
  theme_plots() +
  coord_cartesian(ylim = c(0.3, 0.7))

print(cultural_plot)
```

## 

```{r, echo=FALSE, warning=FALSE, fig.width=12, fig.height=6}

library(ggplot2)
library(dplyr)

theme_plots <- function(base_size = 20, base_family = "Jost Regular") {
  theme_bw(base_size, base_family) +
    theme(
      panel.background = element_rect(fill = "#ffffff", colour = NA),
      title = element_text(size = rel(1), family = "Jost Bold", face = "plain"),
      plot.subtitle = element_text(
        size = rel(0.8),
        family = "Jost Regular",
        face = "plain"
      ),
      plot.caption = element_text(
        margin = margin(t = 10),
        size = rel(0.6),
        family = "Jost Light",
        face = "plain"
      ),
      panel.border = element_rect(
        color = "grey50",
        fill = NA,
        linewidth = 0.15
      ),
      panel.spacing = unit(1, "lines"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.25, colour = "grey90"),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_text(
        size = rel(0.8),
        family = "Jost Medium",
        face = "plain"
      ),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.title.y = element_text(hjust = 1, margin = margin(r = 10)),
      legend.position = "bottom",
      legend.title = element_text(
        size = rel(0.8),
        vjust = 0.5,
        family = "Jost Medium",
        face = "plain"
      ),
      legend.key.size = unit(0.7, "line"),
      legend.key = element_blank(),
      legend.spacing = unit(0.1, "lines"),
      legend.justification = "left",
      legend.margin = margin(t = -5, b = 0, l = 0, r = 0),
      strip.text = element_text(
        size = rel(0.9),
        hjust = 0,
        family = "Jost Regular",
        face = "plain"
      ),
      strip.background = element_rect(fill = "white", colour = NA),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")
    )
}

figure3_data <- tribble(
  ~dimension                , ~congruence     , ~party_type     , ~predicted_support , ~lower_ci , ~upper_ci ,
  # Data for Civil Liberties
  "Civil liberties"         , "Not congruent" , "Populist"      , 0.55               , 0.53      , 0.57      ,
  "Civil liberties"         , "Not congruent" , "Illiberal"     , 0.49               , 0.47      , 0.51      ,
  "Civil liberties"         , "Not congruent" , "Authoritarian" , 0.38               , 0.36      , 0.40      ,
  "Civil liberties"         , "Not congruent" , "Liberal"       , 0.52               , 0.50      , 0.54      ,
  "Civil liberties"         , "Congruent"     , "Populist"      , 0.59               , 0.57      , 0.61      ,
  "Civil liberties"         , "Congruent"     , "Illiberal"     , 0.53               , 0.51      , 0.55      ,
  "Civil liberties"         , "Congruent"     , "Authoritarian" , 0.42               , 0.40      , 0.44      ,
  "Civil liberties"         , "Congruent"     , "Liberal"       , 0.56               , 0.54      , 0.58      ,

  # Data for Rule of Law
  "Rule of law"             , "Not congruent" , "Populist"      , 0.47               , 0.45      , 0.49      ,
  "Rule of law"             , "Not congruent" , "Illiberal"     , 0.48               , 0.46      , 0.50      ,
  "Rule of law"             , "Not congruent" , "Authoritarian" , 0.46               , 0.44      , 0.48      ,
  "Rule of law"             , "Not congruent" , "Liberal"       , 0.53               , 0.51      , 0.55      ,
  "Rule of law"             , "Congruent"     , "Populist"      , 0.51               , 0.49      , 0.53      ,
  "Rule of law"             , "Congruent"     , "Illiberal"     , 0.515              , 0.495     , 0.535     ,
  "Rule of law"             , "Congruent"     , "Authoritarian" , 0.50               , 0.48      , 0.52      ,
  "Rule of law"             , "Congruent"     , "Liberal"       , 0.57               , 0.55      , 0.59      ,

  # Data for Elections
  "Free and fair elections" , "Not congruent" , "Populist"      , 0.55               , 0.53      , 0.57      ,
  "Free and fair elections" , "Not congruent" , "Illiberal"     , 0.42               , 0.40      , 0.44      ,
  "Free and fair elections" , "Not congruent" , "Authoritarian" , 0.41               , 0.39      , 0.43      ,
  "Free and fair elections" , "Not congruent" , "Liberal"       , 0.58               , 0.56      , 0.60      ,
  "Free and fair elections" , "Congruent"     , "Populist"      , 0.58               , 0.56      , 0.60      ,
  "Free and fair elections" , "Congruent"     , "Illiberal"     , 0.45               , 0.43      , 0.47      ,
  "Free and fair elections" , "Congruent"     , "Authoritarian" , 0.44               , 0.42      , 0.46      ,
  "Free and fair elections" , "Congruent"     , "Liberal"       , 0.61               , 0.59      , 0.63
)

party_colors <- c(
  "Authoritarian" = "red",
  "Illiberal" = "purple",
  "Populist" = "blue",
  "Liberal" = "orange"
)

figure3_data$party_type <- factor(
  figure3_data$party_type,
  levels = c("Authoritarian", "Illiberal", "Populist", "Liberal")
)

economic_plot <- ggplot(
  figure3_data,
  aes(x = congruence, y = predicted_support, color = party_type)
) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  geom_errorbar(
    aes(ymin = lower_ci, ymax = upper_ci),
    width = 0.2,
    position = position_dodge(width = 0.5)
  ) +
  facet_wrap(~dimension, ncol = 3) +
  scale_color_manual(values = party_colors) +
  labs(
    title = "Effects of economic policy congruence",
    x = "Economic congruence",
    y = "Predicted support",
    color = "Party type"
  ) +
  theme_plots() +
  coord_cartesian(ylim = c(0.3, 0.7))

print(economic_plot)
```

## Conclusions

-   Citizens distinguish between types of violations; authoritarian most punished
-   Populist violations less punished, sometimes rewarded
-   Electoral fairness violations most consistently punished
-   Policy congruence, especially cultural, reduces willingness to punish even grave violations
-   Conjoint experiments have limits; future research could examine justification strategies
