\documentclass[11pt, a4paper]{article} % Document font size and paper size
\usepackage{fontspec} % Allows the use of OpenType fonts
\usepackage{geometry} % Allows the configuration of document margins
\usepackage[super]{nth}
\geometry{a4paper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in} % Document margin settings
\setlength\parindent{0in} % Remove paragraph indentation
\usepackage[usenames,dvipsnames]{xcolor} % Custom colors
\usepackage{sectsty} % Allows changing the font options for sections in a document
\usepackage[normalem]{ulem} % Custom underlining
\usepackage{xunicode} % Allows generation of unicode characters from accented glyphs
\defaultfontfeatures{Mapping=tex-text} % Converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
\usepackage{marginnote} % For margin years
\newcommand{\years}[1]{\marginnote{\scriptsize #1}} % New command for including margin years
\renewcommand*{\raggedleftmarginnote}{}
\setlength{\marginparsep}{7pt} % Slightly increase the distance of the margin years from the contant
\reversemarginpar
\usepackage[xetex, bookmarks, colorlinks, breaklinks, pdftitle={Benjamin David Stanley - vita},pdfauthor={Benjamin David Stanley}]{hyperref} % PDF setup - set your name and the title of the document to be incorporated into the final PDF file meta-information
\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=MidnightBlue} % Link colors

\setromanfont [Ligatures={Common}, Numbers={OldStyle}]{Hoefler Text} % Main text font
\setmonofont[Scale=0.8]{Monaco} % Set mono font (e.g. phone numbers)
\setsansfont[Scale=0.9]{Hoefler Text} % Set sans font, used in the main name and titles in the document
\newcommand{\amper}{{\fontspec[Scale=.95]{Hoefler Text}\selectfont\itshape\&}} % Custom ampersand character
\sectionfont{\sffamily\mdseries\large\underline} % Set font options for sections
\subsectionfont{\rmfamily\mdseries\upshape\normalsize\emph} % Set font options for subsections
\subsubsectionfont{\rmfamily\bfseries\upshape\normalsize} % Set font options for subsubsections

%----------------------------------------------------------------------------------------

\begin{document}
\textbf{Benjamin David Stanley}
\section*{Qualifications}
\years{2005--2010}\textsc{Ph.D.} in Government, \emph{University of Essex} (ESRC funded)\\`Populism in the Polish party system: party appeals and voter mobilisation’. Supervisor -- Prof. Frances Millard (The University of Essex); internal examiner -- Dr. Sarah Birch; external examiner -- Prof. Aleks Szczerbiak (University of Sussex).\\
\years{2004--2005}\textsc{M.A.} in European Politics, \emph{The University of Essex} (distinction, ESRC funded)\\Thesis title: `The controversial concept of populism'.

%------------------------------------------------
\section*{Employment}

\years{2015--2019}Assistant professor, Centre for the Study of Democracy, SWPS University of Social Sciences and Humanities, Warsaw.\\
\years{2013--2015}Marie Curie Intra-European Research Fellow, School of Law, Politics and Sociology, \emph{University of Sussex}, Brighton.\\
\years{2011--2015}Assistant professor, Institute of Political Science, Faculty of History and Social Sciences, \emph{Cardinal Stefan Wyszyński University}, Warsaw.

%------------------------------------------------
\section*{Selected research projects}
\years{2019-2020}`DIASPOlitic: Understanding the Political Dynamics of Émigré Communities in an Era of European Democratic Backsliding'. Project funded by the Research Council of Norway (€52615). Principal investigator.\\
\years{2013--2015}`The Winner-Loser Divide? A Comparative Analysis of Voting Behaviour and Cleavage Formation in Post-Communist Party Systems' (WINLOSE). Project funded by the Marie Curie Intra-European Fellowship programme, a Marie Curie Action under the Seventh Framework Programme (€231283).

%------------------------------------------------
\section*{Selected publications}
\years{2018}`A New Populist Divide? Correspondences of Supply and Demand in the 2015 Polish Parliamentary Elections.' \emph{East European Politics and Societies}, online publication.\\
\years{2015}`Confrontation by default and confrontation by design: strategic and institutional responses to Poland's populist coalition government’, \emph{Democratization}, 23:2, pp. 263--282, (sole author).\\
\years{2015}`Poland’s Palikot Movement: Voice of the Disenchanted, Missing Ideological Link, or More of the Same?’, \emph{Party Politics}, 22:6, pp. 705--718, (co-authored with Mikołaj Cześnik, corresponding author).\\
\years{2011}`Populism, Nationalism, or National Populism? An Analysis of Slovak Voting Behaviour at the 2010 Parliamentary Elections’, \emph{Communist and Post-Communist Studies}, 44:4, pp. 257--270, (sole author).\\
\years{2008}`The Thin Ideology of Populism’, \emph{Journal of Political Ideologies}, 13:1, pp. 95-110, (sole author).
\end{document}